---
title: "A continuous-wave fNIRS pipeline with synthetic data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A continuous-wave fNIRS pipeline with synthetic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`PhysioNIRS` carries continuous-wave functional near-infrared spectroscopy
(fNIRS) data in a governed `PhysioExperiment` and preserves measurement order,
probe geometry, the exact time base, and processing provenance through every
step. This vignette walks the standard preprocessing-to-analysis path

> raw intensity &rarr; optical density &rarr; haemoglobin concentration (MBLL)
> &rarr; motion correction &rarr; quality control &rarr; short-separation
> regression &rarr; activation GLM

entirely offline on a tiny synthetic recording, so no SNIRF file or device is
required.

```{r setup}
library(PhysioNIRS)
```

## A small synthetic recording

A governed continuous-wave object needs an intensity assay, a `time_seconds`
column, a SNIRF measurement list, and a probe. The recording below has two
source-detector pairs -- one long (3 cm) and one short (0.8 cm) -- each sampled
at 760 nm and 850 nm for 16 s at 10 Hz, with a shared cardiac pulsation, a slow
task response, and one motion spike.

```{r build}
fs <- 10
n_time <- 160L
t <- seq_len(n_time) / fs
cardiac <- 0.06 * sin(2 * pi * 1.0 * t)
task <- 0.08 * exp(-((t - 8)^2) / (2 * 2^2))
base_od <- c(1.2, 1.0, 1.1, 0.9)
wl_gain <- c(1.0, 0.8, 1.0, 0.8)
od_true <- sapply(1:4, function(j) base_od[j] + cardiac + wl_gain[j] * task)
od_true[round(n_time / 2), ] <- od_true[round(n_time / 2), ] + 0.7   # motion
intensity <- sweep(exp(-od_true), 2L, c(5000, 4000, 5200, 4100), "*")
channel_label <- c("S1_D1_760", "S1_D1_850", "S2_D2_760", "S2_D2_850")
colnames(intensity) <- channel_label

measurement <- S4Vectors::DataFrame(
  measurement_index = 1:4,
  source_index = c(1L, 1L, 2L, 2L), detector_index = c(1L, 1L, 2L, 2L),
  wavelength_index = c(1L, 2L, 1L, 2L), wavelength_nm = c(760, 850, 760, 850),
  wavelength_actual_nm = NA_real_, data_type = 1L, data_type_index = 1L,
  data_type_label = NA_character_, data_unit = NA_character_,
  source_power = NA_real_, detector_gain = NA_real_,
  source_label = c("S1", "S1", "S2", "S2"),
  detector_label = c("D1", "D1", "D2", "D2"), channel_label = channel_label)

probe <- list(
  wavelengths = c(760, 850), sourceLabels = c("S1", "S2"),
  detectorLabels = c("D1", "D2"),
  sourcePos2D = rbind(c(0, 0), c(0.1, 0)),
  detectorPos2D = rbind(c(0, 0.03), c(0.1, 0.008)), LengthUnit = "m")

pe <- PhysioExperiment::PhysioExperiment(
  assays = list(raw = intensity),
  rowData = S4Vectors::DataFrame(time_seconds = t),
  colData = cbind(measurement, label = channel_label),
  metadata = list(snirf = list(measurement_list = measurement, probe = probe,
                               stim = list(), time_sampling = "uniform")),
  samplingRate = fs)
pe <- PhysioExperiment::setEvents(pe, PhysioExperiment::PhysioEvents(
  onset = 6, duration = 4, type = "task", value = "task"))
dim(pe)
```

The probe geometry yields the source-detector distance of each measurement:

```{r distances}
sourceDetectorDistances(pe)
```

## Optical density

`intensityToOD()` applies the natural-log convention `-log(I / I0)` and records
the reference it used in provenance.

```{r od}
od <- intensityToOD(pe)
SummarizedExperiment::assayNames(od)
```

## Haemoglobin concentration (MBLL)

`mbll()` collapses each source-detector pair's two wavelengths into HbO, HbR,
and HbT concentrations using the modified Beer-Lambert law, deriving the
differential pathlength from the probe geometry.

```{r mbll}
hb <- mbll(od)
SummarizedExperiment::assayNames(hb)
dim(hb)
```

## Motion correction

Motion shows up as abrupt shifts in optical density. Detect it, then correct it
with temporal-derivative distribution repair (`tddr()`); spline- and
wavelet-based correctors are also available.

```{r motion}
mask <- motionArtifactDetect(od)
corrected <- tddr(od)
SummarizedExperiment::assayNames(corrected)
```

## Signal quality and channel pruning

The scalp-coupling index and a peak-power signal-quality index both score
whether a channel carries a clean cardiac pulsation. `pruneChannels()` then
marks or drops channels that fail.

```{r quality}
sqi <- signalQualityIndex(od)
sqi
pruned <- pruneChannels(od, sqi, action = "mark")
dim(pruned)
```

## Short-separation regression

Short channels mostly see scalp haemodynamics. Identify them by distance, then
regress their signal out of the long channels as a systemic-physiology
nuisance.

```{r short}
short <- identifyShortChannels(od, threshold_m = 0.015)
short[, c("channel_id", "distance_m", "is_short")]
od_ssr <- shortSeparationRegress(od, short = short)
SummarizedExperiment::assayNames(od_ssr)
```

## Activation GLM

Finally, fit a general linear model of the HbO concentration against an HRF
regressor built from the object's stimulus events, and read off a per-channel
task contrast.

```{r glm}
fit <- nirsActivationGLM(hb, assay_name = "HbO")
fit$coefficients
nirsActivationContrast(fit, c(task = 1))
```

## Where to go next

* **Real data.** `readSNIRF()` and `writeSNIRF()` move governed objects to and
  from SNIRF/HDF5 files, preserving the measurement list, probe, stimulus
  tables, and provenance.
* **Live neurofeedback.** `nirsNeurofeedback()` and its lifecycle functions run
  a closed loop over a regular-rate HbO stream; they require a live stream (for
  example an LSL inlet provided by `PhysioStream`) and so are not exercised here.
* **Upstream and downstream packages.** `PhysioExperiment` defines the shared
  data model, and `PhysioStream` provides the streaming transports used by the
  neurofeedback loop.
```
